---
title: "ggplot2"
author: "Pierre Gestraud"
date: "4 octobre 2016"
output: 
  html_notebook: default
  github_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
knitr::opts_knit$set(cache = TRUE)
require(ggplot2)
```

### Example data

```{r}
head(mpg)
```

### Quickstart

```{r}
ggplot(mpg, aes(x = displ, y = hwy)) + geom_point()
```

A plot has 3 main components:

- data
- a set of aesthetics
- a set of layers (mainly geometries)

```{r}
ggplot(mpg,   ## data
       aes(x = displ, y = hwy)) +  ## aesthetics
    geom_point() ## layer
```

#### add geoms
```{r}
ggplot(mpg, aes(x = displ, y = hwy)) + geom_point() + geom_line()
```


### aesthetics

- `x`
- `y`
- `z` 
- `color`
- `fill`
- `size`
- `text`
- `label`
- `shape`
- `linetype`
- `alpha`
- `group`

Set color aesthetic to `manufacturer variable`

```{r}
ggplot(mpg, aes(x = displ, y = hwy, color = manufacturer)) + geom_point()
```

Set size aesthetic

```{r}
ggplot(mpg, aes(x = displ, y = hwy, size = cyl)) + geom_point()
```


Alpha aesthetic

```{r}
ggplot(mpg, aes(x = displ, y = hwy, size = cyl)) + geom_point(alpha = 0.5)
```


Set an aesthetic to a fixed value

```{r}
ggplot(mpg, aes(x = displ, y = hwy)) + geom_point(aes(color = "blue"))
ggplot(mpg, aes(x = displ, y = hwy)) + geom_point(color = "blue")
```


Set an aestetic to a discrete numeric value
```{r}
ggplot(mpg, aes(x = displ, y = hwy, color = cyl)) + geom_point() ## gradient scale
ggplot(mpg, aes(x = displ, y = hwy, color = factor(cyl))) + geom_point() ## discrete scale
```

Aesthetics can be defined inside geom

```{r}
ggplot(mpg, aes(x = displ, y = hwy)) + geom_point(aes(color = class))
```

Aesthetics defined inside `ggplot` apply to all geom

```{r}
ggplot(mpg, aes(x = class, y = hwy, color = class)) + geom_boxplot() + geom_point()
ggplot(mpg, aes(x = class, y = hwy)) + geom_boxplot(aes(color = class)) + geom_point()
```

### a ggplot is an object

We can build a `ggplot` object in several steps

```{r}
g <- ggplot(mpg, aes(x = displ, y = hwy)) + geom_point() 
g ## or print(g)
g <- g + ggtitle("hwy vs displ")
g
```

### geom

* one variable
    + continous
        - `geom_density`
        - `geom_histogram`
        - `geom_freqpoly`
    + discrete
        - `geom_bar`
* two variables
    + continous X, continuous Y
        - `geom_point`
        - `geom_smooth`
        - `geom_text`
        - `geom_line`
    + discrete X, continuous Y
        - `geom_bar`
        - `geom_boxplot`
        - `geom_violin`
    + bivariate distribution
        - `geom_density2d`
        - `geom_hex`
        - `geom_bin2d`
    + continuous function
        - `geom_line`
        - `geom_step` 
* three variables
    +
        - `geom_contour`
        - `geom_tile`


#### data distributions

For histograms and densities only `x` is mandatory

```{r}
ggplot(mpg, aes(x = hwy)) + geom_histogram()
ggplot(mpg, aes(x = hwy)) + geom_density()
ggplot(mpg, aes(x = hwy)) + geom_freqpoly()
```

Changing bins

```{r}
ggplot(mpg, aes(x = hwy)) + geom_histogram(bins = 30)
ggplot(mpg, aes(x = hwy)) + geom_histogram(binwidth = 2)
```

Manually setting breaks

```{r}
ggplot(mpg, aes(x = hwy)) + geom_histogram(breaks = c(10, 15, 20, 25, 30, 40, 50))
```

Add fill or color

```{r}
ggplot(mpg, aes(x = hwy)) + geom_density(fill = "blue", color = "red")
ggplot(mpg, aes(x = hwy)) + geom_density(aes(fill = class))
```


#### Boxplots, violins, jitter

boxplots need `x` and `y` aestetics

```{r}
ggplot(mpg, aes(x = class, y = hwy)) + geom_boxplot()
ggplot(mpg, aes(x = class, y = hwy, fill = class)) + geom_boxplot()
```

Add another variable as fill or color will separate boxplots

```{r}
ggplot(mpg, aes(x = class, y = hwy, fill = trans)) + geom_boxplot()
```

Violin plots

```{r}
ggplot(mpg, aes(x = class, y = hwy)) + geom_violin()
```

Jitters

```{r}
ggplot(mpg, aes(x = class, y = hwy)) + geom_jitter()
## control width and height of jitter
ggplot(mpg, aes(x = class, y = hwy)) + geom_jitter(width = 0.5, height = 0)
```

```{r}
ggplot(mpg, aes(x = class, y = hwy)) + geom_boxplot() + geom_jitter(width = 0.5, height = 0)
```

#### Bars

```{r}
ggplot(mpg, aes(x = manufacturer)) + geom_bar()
```

Adding color

```{r}
ggplot(mpg, aes(x = manufacturer, fill = class)) + geom_bar()
```

Changing bars positions

```{r}
ggplot(mpg, aes(x = manufacturer, fill = class)) + geom_bar(position = "stack")
ggplot(mpg, aes(x = manufacturer, fill = class)) + geom_bar(position = "dodge")
ggplot(mpg, aes(x = manufacturer, fill = class)) + geom_bar(position = "fill")
```

Bar plot from count data

```{r}
d <- data.frame(group = c("a", "b", "c"), n = c(10, 25 , 14))
d
```

```{r}
## error
## ggplot(d, aes(x = group, y = n)) + geom_bar()
ggplot(d, aes(x = group, y = n)) + geom_bar(stat = "identity")
## or
ggplot(d, aes(x = group, y = n)) + geom_col()
```


### add a smoother on the plot
Default smoother (loess if <1000 obs, gam otherwise)

```{r}
ggplot(mpg, aes(x = displ, y = hwy)) + geom_point() + geom_smooth()
```

Linear model

```{r}
ggplot(mpg, aes(x = displ, y = hwy)) + geom_point() + geom_smooth(method = "lm")
```

Remove SE

```{r}
ggplot(mpg, aes(x = displ, y = hwy)) + geom_point() + geom_smooth(method = "lm", se = FALSE)
```

Smooth by group

```{r}
ggplot(mpg, aes(x = displ, y = hwy, color = class)) + geom_point() + geom_smooth(method = "lm", se = FALSE)
```


### Order of layers in object construction

Order of layers matters

```{r}
ggplot(mpg, aes(x = class, y = hwy)) + geom_jitter(width = 0.5, height = 0) + geom_boxplot() 
ggplot(mpg, aes(x = class, y = hwy)) + geom_boxplot() + geom_jitter(width = 0.5, height = 0) 
```


### Scales

Scales are mainly controlled by `scale_*` functions. Their names are structured: *scale_'aestetic name'_'type'*

* `scale_color_discrete`: discrete scale for color aestetic
* `scale_fill_gradient`: continuous scale for fill aestetic


#### Discrete scales

```{r}
ggplot(mpg, aes(x = displ, y = hwy, color = class)) + geom_point() + scale_color_discrete(h = c(20, 150))
ggplot(mpg, aes(x = displ, y = hwy, color = class)) + geom_point() + scale_color_brewer(palette = "Set2")
ggplot(mpg, aes(x = displ, y = hwy, color = factor(cyl))) + geom_point() + scale_color_manual(values = c("blue", "black", "red", "pink"))
ggplot(mpg, aes(x = displ, y = hwy, color = factor(cyl))) + geom_point() + scale_color_manual(values = c('8' = "blue", '6' = "black", '4' = "red", '5' = "pink"))
```

#### gradient scales

```{r}
ggplot(mpg, aes(x = displ, y = hwy, color = hwy)) + geom_point()

## 2 colors gradient
ggplot(mpg, aes(x = displ, y = hwy, color = hwy)) + geom_point() + scale_color_gradient(low = "green", high = "red")

## diverging color gradient, define low, mid and high colors
ggplot(mpg, aes(x = displ, y = hwy, color = hwy)) + geom_point() + scale_color_gradient2(low = "green", high = "red", mid = "black", midpoint = 25)

## n colors gradient
ggplot(mpg, aes(x = displ, y = hwy, color = hwy)) + geom_point() + scale_color_gradientn(colours = terrain.colors(10))
```


#### `xlim`/`ylim` vs `coord_cartesian`

* `xlim`/`ylim` select data inside the range
* coord_cartesian zoom in the plot

```{r}
ggplot(mpg, aes(x = displ, y = hwy)) + geom_point()
ggplot(mpg, aes(x = displ, y = hwy)) + geom_point() + xlim(c(2, 6))
ggplot(mpg, aes(x = displ, y = hwy)) + geom_point() + coord_cartesian(xlim = c(2, 6))
```

Warning: when computing distributions or smoothers, the plot will completely change by using `xlim`/`ylim`

```{r}
ggplot(mpg, aes(x = displ)) + geom_density(fill = "gray50")
ggplot(mpg, aes(x = displ)) + geom_density(fill = "gray50") + xlim(c(2, 3))
ggplot(mpg, aes(x = displ)) + geom_density(fill = "gray50") + coord_cartesian(xlim = c(2, 3))
```

#### axis transformation

* flip axes

```{r}
ggplot(mpg, aes(x = displ, y = hwy)) + geom_point() + coord_flip()
ggplot(mpg, aes(x = class, y = displ)) + geom_boxplot() + coord_flip()
```

* polar coordinates

```{r}
ggplot(mpg, aes(x = displ, y = hwy)) + geom_point() + coord_polar()
ggplot(mpg, aes(x = class, y = displ)) + geom_boxplot() + coord_polar()
```

* reverse axis

```{r}
ggplot(mpg, aes(x = displ, y = hwy)) + geom_point() + scale_x_reverse()
```

* log axis

```{r}
ggplot(mpg, aes(x = displ, y = hwy)) + geom_point() + scale_x_log10() 
```

* order of discrete scales

```{r}
ggplot(mpg, aes(x = fl)) + geom_bar()

## change levels order 
mpg$fl2 <- factor(mpg$fl, levels = c("e", "p", "d", "r", "c"))
ggplot(mpg, aes(x = fl2)) + geom_bar()

## or use xlim
ggplot(mpg, aes(x = fl)) + geom_bar() + xlim(c("e", "p", "d", "r", "c"))
```


### Themes

Themes allow to have a set of graphics with the same style. Some are predefined `theme_default, theme_bw, them_minimal, theme_gray...`

```{r}
ggplot(mpg, aes(x = displ, y = hwy)) + geom_point() + theme_bw()
ggplot(mpg, aes(x = displ, y = hwy)) + geom_point() + theme_minimal()
ggplot(mpg, aes(x = displ, y = hwy)) + geom_point() + theme_dark()
```

We can customize some elements in themes (see http://docs.ggplot2.org/dev/vignettes/themes.html for details)

```{r}
ggplot(mpg, aes(x = displ, y = hwy)) + geom_point() + theme(panel.background = element_rect(fill = "wheat"), axis.text = element_text(size = 20))
```

User defined theme

```{r}
my_theme <- function() theme_bw() %+replace% theme(panel.background = element_rect(fill = "wheat"))
g
g + my_theme()
```


### Faceting

Facetting allows to display plot separately according to discrete variables

Two functions for facetting:
- `facet_wrap`
- `facet_grid`

```{r}
ggplot(mpg, aes(x = displ, y = hwy)) + geom_point() + facet_wrap(~class)
ggplot(mpg, aes(x = displ, y = hwy)) + geom_point() + facet_grid(.~class)
ggplot(mpg, aes(x = displ, y = hwy)) + geom_point() + facet_grid(class~drv)
```

Controling facet_wrap organisation

```{r}
ggplot(mpg, aes(x = displ, y = hwy)) + geom_point() + facet_wrap(~class, ncol = 2)
ggplot(mpg, aes(x = displ, y = hwy)) + geom_point() + facet_wrap(~class, scales = "free")
ggplot(mpg, aes(x = displ, y = hwy)) + geom_point() + facet_wrap(~class, scales = "free_x")
```

All layers are faceted

```{r}
ggplot(mpg, aes(x = displ, y = hwy)) + geom_point() + geom_smooth() + facet_grid(~drv)
```


### add supplementary information on plot

#### add lines
```{r}
ggplot(mpg, aes(x = displ, y = hwy)) + geom_point() + geom_hline(yintercept = 20)
ggplot(mpg, aes(x = displ, y = hwy)) + geom_point() + geom_vline(xintercept = 4.5)
ggplot(mpg, aes(x = displ, y = hwy)) + geom_point() + geom_abline(slope = -2, intercept = 35)
```

#### annotate

```{r}
ggplot(mpg, aes(x = displ, y = hwy)) + geom_point() + annotate(x = 4, y = 30, geom = "point", color = "red", size = 4)
ggplot(mpg, aes(x = displ, y = hwy)) + geom_point() + annotate(x = 4, y = 30, geom = "label", label = "annotation") 
```

#### add data from external data.frame
```{r}
d_annot <- data.frame(xpos = c(2, 5, 8), ypos = c(20, 25, 40), lab = c("lab1", "lab2", "lab3"))
ggplot(mpg, aes(x = displ, y = hwy)) + geom_point() + geom_label(aes(x = xpos, y = ypos, label = lab), data = d_annot)
```


### Programming with ggplot2

`ggplot2` use non standard evaluation (NSE). How to use variables in `aes`? `aes_string` allows standard evaluation.

```{r}
my_var <- "displ"
ggplot(mpg, aes(x = my_var, y = hwy)) + geom_point()
ggplot(mpg, aes_string(x = my_var, y = "hwy")) + geom_point()
```


### Interactivity
For interactive graphics see `ggvis` or `plotly` packages

```{r, eval = FALSE}
require(ggvis)
ggvis(mpg, ~displ, ~hwy) %>% layer_points()
```
```{r, eval=FALSE}
ggvis(mpg, ~displ, ~hwy) %>%  
    layer_smooths(span = input_slider(0.2, 1, value = 1), stroke := "blue") %>% 
    layer_points(size := input_slider(10, 200, value = 50))
```


```{r, eval = FALSE}
require(plotly)
g <- ggplot(mpg, aes(x = displ, y = hwy)) + geom_point()
ggplotly(g)
```


### Extensions

- `ggrepel`: smart geom_text and geom_label placement
- `cowplot`: several plots on one page
- `ggraph`: visualise graph
- `ggmap`: plot data on a map
- `factoextra`: visualise results of factorial analysis (PCA, CA, MCA, MFA, ...)
- `ggbio`: visualise genomic data
- `ggdendro`: visualise dendrograms and trees
- `ggthemes`: additionnal themes

### Ressources

- http://docs.ggplot2.org
- rstudio cheatsheet http://www.rstudio.com/wp-content/uploads/2015/04/ggplot2-cheatsheet.pdf
- ggplot2 book https://github.com/hadley/ggplot2-book