---
title: "ggplot2"
author: "Pierre Gestraud"
date: "4 octobre 2016"
output: 
  html_document: 
    self_contained: no
    keep_md: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
knitr::opts_knit$set(cache = TRUE)
require(ggplot2)
```


### Example data

```{r}
head(msleep)
```

Create additionnal variable

```{r}
msleep$continent <- rep(c("Europe", "Africa", "Asia"), length.out = nrow(msleep))
```


### First plot

```{r}
qplot(sleep_total, sleep_rem, data = msleep)
```

```{r}
qplot(sleep_total, sleep_rem, color = vore, data = msleep)
```

```{r}
qplot(sleep_total, sleep_rem, size = bodywt, data = msleep)
```

### Advanced control 

main function: `ggplot` 

### Grammar of graphics

plot = data + geom + coordinate system

### Layer

a layer is composed of 

- data and aesthetic mapping,
- a statistical transformation,
- a geometric object
- and a position adjustement

### 

```{r}
ggplot(msleep,  ## data
       aes(x = sleep_total, y = sleep_rem)) + ## aesthetics
        geom_point() + ## layer
        ggtitle("REM vs total") ## additionnal elements
    ```

### a plot is an R object

```{r}
g <- ggplot(msleep, aes(x = sleep_total, y = sleep_cycle)) + geom_point() 
g ## or print(g)
g + ggtitle("cycle vs total")
```


### geom

* one variable
    + continous
        - `geom_density`
        - `geom_histogram`
        - `geom_freqpoly`
    + discrete
        - `geom_bar`
* two variables
    + continous X, continuous Y
        - `geom_point`
        - `geom_smooth`
        - `geom_text`
    + discrete X, continuous Y
        - `geom_bar`
        - `geom_boxplot`
        - `geom_violin`
    + bivariate distribution
        - `geom_density2d`
        - `geom_hex`
        - `geom_bin2d`
    + continuous function
        - `geom_line`
        - `geom_step`
* three variables
    +
        - `geom_contour`
        - `geom_tile`

```{r}
ggplot(msleep, aes(x = vore, y = sleep_total)) + geom_boxplot()
ggplot(msleep, aes(x = vore)) + geom_bar()
```


### aesthetics

- `x`
- `y`
- `z` 
- `color`
- `fill`
- `size`
- `text`

```{r}
ggplot(msleep, aes(x = vore, y = sleep_rem, fill = vore)) + geom_boxplot()
ggplot(msleep, aes(x = vore, y = sleep_rem, color = vore)) + geom_boxplot()
ggplot(msleep, aes(x = vore, y = sleep_rem, fill = continent)) + geom_boxplot()
```


### add a smoother on the plot
Default smoother (loess if <1000 obs, gam otherwise)
```{r}
ggplot(msleep, aes(x = sleep_total, y = sleep_cycle)) + geom_point() + geom_smooth()
```

Linear model
```{r}
ggplot(msleep, aes(x = sleep_total, y = sleep_cycle)) + geom_point() + geom_smooth(method = "lm")
```

Remove SE
```{r}
ggplot(msleep, aes(x = sleep_total, y = sleep_cycle)) + geom_point() + geom_smooth(method = "lm", se = FALSE)
```

### Order matters
```{r}
ggplot(msleep, aes(x = vore, y = sleep_total)) + geom_point() + geom_boxplot()
ggplot(msleep, aes(x = vore, y = sleep_total)) + geom_boxplot() + geom_point()
```

### Scales

- `xlim`/`ylim` vs `coord_cartesian`

```{r}
ggplot(msleep, aes(x = sleep_total, y = sleep_cycle)) + geom_point() + xlim(c(7.5, 17))
ggplot(msleep, aes(x = sleep_total, y = sleep_cycle)) + geom_point() + coord_cartesian(xlim = c(7.5, 17))

ggplot(msleep, aes(x = sleep_total)) + geom_density(fill = "gray") 
ggplot(msleep, aes(x = sleep_total)) + geom_density(fill = "gray") + xlim(c(7.5, 17))
ggplot(msleep, aes(x = sleep_total)) + geom_density(fill = "gray") + coord_cartesian(xlim = c(7.5, 17))
```



- axis transformation
- order
- color scales

### Themes

- default themes
- customize theme

### Faceting

Two functions for facetting:
- `facet_wrap`
- `facet_grid`

```{r}
ggplot(msleep, aes(x = sleep_total, y = sleep_rem)) + geom_point() + facet_wrap(~ vore)
ggplot(msleep, aes(x = sleep_total, y = sleep_rem)) + geom_point() + facet_grid(continent ~ vore)
```

Controling facet_wrap organisation

```{r}
ggplot(msleep, aes(x = sleep_total, y = sleep_rem)) + geom_point() + facet_wrap(~ vore, nrow = 1)
```

All layers are faceted
```{r}
ggplot(msleep, aes(x = sleep_total, y = sleep_rem)) + geom_point() + facet_wrap(~ vore) + geom_smooth(se = FALSE, method = "lm")
```


- scales free
- labels

### add supplementary data

### Programming with ggplot2

`ggplot2` use non standard evaluation (NSE). How to use variables in `aes`? `aes_string` allows standard evaluation.

```{r}
var_x <- "sleep_total"
var_y <- "sleep_rem"
ggplot(msleep, aes(x = var_x, y = var_y)) + geom_point()
ggplot(msleep, aes_string(x = var_x, y = var_y)) + geom_point()
```


### Plot heatmap with ggplot2

```{r}
ggplot(msleep, aes(x = continent, y = vore, fill = sleep_total)) + geom_tile()
```

```{r}
d <- as.data.frame(matrix(rnorm(100), ncol = 10))
colnames(d) <- paste0("s", 1:10)
d$gene <- paste0("g", 1:10)
require(tidyr)
dm <- gather(d, key = "sample", value = "exp", -gene) ## convert data to long format
ggplot(dm, aes(x = sample, y = gene, fill = exp)) + geom_tile()
```


### Interactivity

```{r}
require(plotly)
g <- ggplot(msleep, aes(x = sleep_total, y = sleep_rem)) + geom_point()
ggplotly(g)
g <- ggplot(msleep, aes(x = sleep_total, y = sleep_rem, color = vore, text = name)) + geom_point()
ggplotly(g)
```


### Extensions

- `ggrepel`: smart geom_text and geom_label placement
- `cowplot`: several plots on one page
- `ggraph`: visualise graph
- `ggmap`: plot data on a map
- `factoextra`: visualise results of factorial analysis (PCA, CA, MCA, MFA, ...)
- `ggbio`: visualise genomic data
- `ggdendro`: visualise dendrograms and trees

### Ressources

- http://docs.ggplot2.org
- rstudio cheatsheet http://www.rstudio.com/wp-content/uploads/2015/04/ggplot2-cheatsheet.pdf